---
title: "Daphnia reciprocal transplant experiment"
---

## Introduction

This document analyses the Daphnia reciprocal transplant experiment, examining relative growth rates and variance partitioning across evolutionary origin, water source, and host environment.

**Note:** This R Markdown document generates:
Mean relative growth rates across treatments
Variance partitioning with bootstrap 95% confidence intervals

### Load libraries

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lme4)
library(plotrix)
library(readxl)
library(patchwork)
library(parallel)
library(lmerTest)
```

### Data import

```{r data-import, message=FALSE, warning=FALSE}
df <- read.table("data_daphnia.txt", sep = "\t", header = TRUE)
```

### Data preparation

```{r data-preparation, message=FALSE, warning=FALSE}
df$daphnia_origin <- factor(df$daphnia_origin, levels = c("control", "heated"))
df$pond_host      <- factor(df$pond_host,      levels = c("control", "heated"))
df$water_origin   <- factor(df$water_origin,   levels = c("control", "heated"))

df$block          <- factor(df$block)
df$pond_recipient <- factor(df$pond_recipient)

options(contrasts = c("contr.sum", "contr.poly"))

df$rel.count <- (log(df$daphnia_end + 1) - log(df$daphnia_start + 1)) / 14
```

### Summary statistics

```{r summary-statistics, message=FALSE, warning=FALSE}
mdata.plot <- df %>%
  group_by(daphnia_origin, pond_host, water_origin) %>%
  summarise(
    fitness_av = mean(rel.count, na.rm = TRUE),
    fitness_se = std.error(rel.count, na.rm = TRUE),
    .groups = "drop"
  )
```

### Mean relative growth rates

```{r panel-a-means, message=FALSE, warning=FALSE, fig.width=5.2, fig.height=4}
px_means <- ggplot(
  mdata.plot,
  aes(
    x = pond_host, y = fitness_av,
    color = daphnia_origin,
    shape = water_origin,
    linetype = water_origin,
    group = interaction(daphnia_origin, water_origin)
  )
) +
  geom_errorbar(
    aes(ymin = fitness_av - fitness_se, ymax = fitness_av + fitness_se),
    width = 0.18, linewidth = 0.7,
    position = position_dodge(0.28)
  ) +
  geom_point(size = 3, position = position_dodge(0.28)) +
  geom_line(linewidth = 1, position = position_dodge(0.28)) +
  scale_color_manual(
    name = "Origin treatment",
    values = c("control" = "#0072B2", "heated" = "#D55E00"),
    labels = c("Control origin", "Heated origin")
  ) +
  scale_shape_manual(
    name = "Water source",
    values = c(control = 16, heated = 17),
    labels = c("Control water", "Heated water")
  ) +
  scale_linetype_manual(
    name = "Water source",
    values = c(control = "solid", heated = "dashed"),
    labels = c("Control water", "Heated water")
  ) +
  labs(
    x = "Host environment",
    y = "Relative growth rate (day\u207B\u00B9)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "top"
  )

print(px_means)
```

### Mixed model and Type III ANOVA

```{r mixed-model, message=FALSE, warning=FALSE}
model <- lmerTest::lmer(rel.count ~ daphnia_origin * pond_host * water_origin + (1 | pond_recipient) + (1 | block), data = df, REML = TRUE)

anova_type3 <- anova(model, type = 3)
print(anova_type3)

anova_df <- as.data.frame(anova_type3)
write.table(anova_df, "anova_typeIII_lmerTest.txt", sep = "\t", quote = FALSE)

if (!("Sum Sq" %in% colnames(anova_df))) {
  stop("No 'Sum Sq' column found in lmerTest::anova output.")
}
```

### Variance partitioning (Type III SS proportions)

```{r variance-partitioning, message=FALSE, warning=FALSE}
label_terms <- function(term_vec) {
  dplyr::recode(
    term_vec,
    "daphnia_origin" = "Evolution (Origin)",
    "water_origin" = "Water-mediated environment",
    "pond_host" = "Environment (Host)",
    "daphnia_origin:water_origin" = "Evo × Water",
    "pond_host:water_origin" = "Env × Water",
    "daphnia_origin:pond_host" = "Evo × Env",
    "daphnia_origin:pond_host:water_origin" = "Evo × Env × Water",
    .default = term_vec
  )
}
desired_order <- c(
  "Water-mediated environment",
  "Evolution (Origin)",
  "Environment (Host)",
  "Evo × Water",
  "Env × Water",
  "Evo × Env",
  "Evo × Env × Water"
)

vp <- data.frame(
  term = rownames(anova_df),
  sumsq = anova_df$`Sum Sq`,
  stringsAsFactors = FALSE
) %>%
  filter(term != "(Intercept)") %>%
  filter(!is.na(sumsq)) %>%
  mutate(
    proportion = sumsq / sum(sumsq),
    term_label = label_terms(term),
    term_label = factor(term_label, levels = desired_order)
  ) %>%
  arrange(term_label)

print(vp)
write.table(vp, "variance_partitioning_typeIII_SS.txt", sep = "\t", quote = FALSE, row.names = FALSE)
```

### Bootstrap confidence intervals

```{r bootstrap-ci, message=FALSE, warning=FALSE}
DO_BOOT <- TRUE

if (DO_BOOT) {
  
  ss_prop_from_anova <- function(anova_df_in) {
    data.frame(
      term = rownames(anova_df_in),
      sumsq = anova_df_in$`Sum Sq`,
      stringsAsFactors = FALSE
    ) %>%
      dplyr::filter(term != "(Intercept)") %>%
      dplyr::filter(!is.na(sumsq)) %>%
      dplyr::mutate(prop = sumsq / sum(sumsq))
  }
  
  vp0 <- ss_prop_from_anova(anova_df)
  
  set.seed(123)
  B <- 300
  
  pond_levels <- levels(df$pond_recipient)
  n_pond <- length(pond_levels)
  
  boot_once <- function(b) {
    sampled_ponds <- sample(pond_levels, size = n_pond, replace = TRUE)
    
    boot_list <- lapply(seq_along(sampled_ponds), function(i) {
      sub <- df[df$pond_recipient == sampled_ponds[i], , drop = FALSE]
      sub$pond_recipient_boot <- factor(paste0("pondDraw", i))
      sub
    })
    boot_df <- do.call(rbind, boot_list)
    
    fit <- tryCatch(
      lmerTest::lmer(
        rel.count ~ daphnia_origin * pond_host * water_origin +
          (1 | pond_recipient_boot) + (1 | block),
        data = boot_df,
        REML = TRUE,
        control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
      ),
      error = function(e) NULL
    )
    if (is.null(fit)) return(NULL)
    
    an <- tryCatch(anova(fit, type = 3), error = function(e) NULL)
    if (is.null(an)) return(NULL)
    
    an_df_b <- as.data.frame(an)
    if (!("Sum Sq" %in% colnames(an_df_b))) return(NULL)
    
    vp_b <- ss_prop_from_anova(an_df_b)
    setNames(vp_b$prop, vp_b$term)
  }
  
  message("Running bootstrap ...")
  boot_res <- lapply(1:B, boot_once)
  boot_res <- boot_res[!vapply(boot_res, is.null, logical(1))]
  message("Successful bootstrap fits: ", length(boot_res), " / ", B)
  
  if (length(boot_res) < 50) stop("Too many bootstrap failures.")
  
  # list -> matrix aligned by term
  all_terms <- sort(unique(unlist(lapply(boot_res, names))))
  boot_mat <- matrix(NA_real_, nrow = length(boot_res), ncol = length(all_terms),
                     dimnames = list(NULL, all_terms))
  for (i in seq_along(boot_res)) boot_mat[i, names(boot_res[[i]])] <- boot_res[[i]]
  
  boot_mean <- apply(boot_mat, 2, function(x) mean(x, na.rm = TRUE))
  
  # build CI table
  vp_ci <- data.frame(
    term = colnames(boot_mat),
    estimate = vp0$prop[match(colnames(boot_mat), vp0$term)],
    mean_boot = boot_mean,
    CI_lower = apply(boot_mat, 2, function(x) quantile(x, 0.025, na.rm = TRUE)),
    CI_upper = apply(boot_mat, 2, function(x) quantile(x, 0.975, na.rm = TRUE)),
    stringsAsFactors = FALSE
  )
  
  vp_ci$estimate[is.na(vp_ci$estimate)] <- vp_ci$mean_boot[is.na(vp_ci$estimate)]
  
  write.table(vp_ci, "variance_partitioning_typeIII_SS_95CI.txt",
              sep = "\t", quote = FALSE, row.names = FALSE)
}
```

### Variance partitioning plot

```{r panel-b-variance, message=FALSE, warning=FALSE, fig.width=5.2, fig.height=4.2}
if (DO_BOOT) {
  
  vp_ci$term_label <- dplyr::recode(
    vp_ci$term,
    "daphnia_origin" = "Evolution",
    "water_origin" = "Water",
    "pond_host" = "Environment",
    "daphnia_origin:water_origin" = "Evo × Water",
    "pond_host:water_origin" = "Env × Water",
    "daphnia_origin:pond_host" = "Evo × Env",
    "daphnia_origin:pond_host:water_origin" = "Evo × Env × Water",
    .default = vp_ci$term
  )
  
  wanted_levels <- c(
    "Water", "Evolution", "Environment",
    "Evo × Water", "Env × Water", "Evo × Env", "Evo × Env × Water"
  )
  
  vp_ci <- vp_ci %>% dplyr::filter(term_label %in% wanted_levels)
  vp_ci$term_label <- factor(vp_ci$term_label, levels = wanted_levels)
  
  vp_ci$color_group <- dplyr::case_when(
    vp_ci$term_label == "Evolution"   ~ "Evolution",
    vp_ci$term_label == "Water"       ~ "Water",
    vp_ci$term_label == "Environment" ~ "Environment",
    TRUE                              ~ "Interaction"
  )
  
  eco_palette <- c(
    "Evolution"   = "#E69F00", 
    "Water"       = "#0F766E",
    "Environment" = "#6B8E23",
    "Interaction" = "#D9D9D9"
  )
  
  px_vp_ci <- ggplot(vp_ci, aes(x = term_label, y = estimate, fill = color_group)) +
    geom_col(width = 0.75, color = "grey20", linewidth = 0.4) +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.18, linewidth = 0.8) +
    scale_fill_manual(values = eco_palette) +
    theme_classic(base_size = 16) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 25, hjust = 1),
      axis.title.y = element_text(face = "bold")
    ) +
    labs(x = NULL, y = "Contribution to variance")
  
  print(px_vp_ci)

}
```
