---
title: "HOBO temperature and light analysis"
---

## Introduction

This document analyses HOBO temperature logger data, examining daily maximum and mean temperatures, mean light across treatments and depths, including effect size calculations and outside logger data.

**Note:** This R Markdown document generates:
- Daily maximum temperature plots with heatwave period and snail sublethal thermal stress 
- Effect size plots (Heated − Control) for daily mean and maximum temperatures 
- Daily mean temperature and mean light plots including outside loggers 

### Load libraries

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(plotrix)
library(patchwork)
```

### Configuration

```{r configuration, message=FALSE, warning=FALSE}
# File paths
temp_file  <- "hobo.xlsx"
temp_sheet <- "All"
tz_local   <- "Europe/Berlin"

# Pond → treatment map
pond_map <- tibble::tribble(
  ~unit_id, ~treatment,
  "1A","control","1B","heated",
  "2A","control","2B","heated",
  "3A","heated","3B","control",
  "4A","heated","4B","control",
  "5A","control","5B","heated",
  "6A","control","6B","heated",
  "7A","heated","7B","control",
  "8A","heated","8B","control",
  "9A","control","9B","heated",
  "10A","control","10B","heated",
  "11A","heated","11B","control",
  "12A","heated","12B","control",
  "13A","control","13B","heated",
  "14A","control","14B","heated",
  "15A","heated","15B","control",
  "16A","heated","16B","control"
)
```

### Plot settings

```{r plot-settings, message=FALSE, warning=FALSE}
theme_pub <- theme_classic(base_size = 15) +
  theme(
    axis.line = element_line(color = "black", linewidth = 0.6),
    strip.background = element_blank(),
    legend.position = "top",
    axis.text = element_text(color = "black"),
    panel.spacing = grid::unit(0.35, "lines")
  )

cols <- c(control = "#0072B2", heated = "#D55E00")
heatwave_col <- "#7E0000"
```

### Data import and processing

```{r data-import, message=FALSE, warning=FALSE}
raw <- read_excel(temp_file, sheet = temp_sheet, col_names = FALSE)

h1 <- as.character(unlist(raw[1, ], use.names = FALSE))
h2 <- as.character(unlist(raw[2, ], use.names = FALSE))
h3 <- as.character(unlist(raw[3, ], use.names = FALSE))
h4 <- as.character(unlist(raw[4, ], use.names = FALSE))
h1[is.na(h1)] <- ""; h2[is.na(h2)] <- ""; h3[is.na(h3)] <- ""; h4[is.na(h4)] <- ""

new_names <- str_replace_all(str_trim(paste(h1, h2, h3, h4, sep = "_")), "\\s+", "")
new_names <- str_replace_all(new_names, "__+", "_")
new_names[new_names == "___" | new_names == ""] <- paste0("V", which(new_names == "___" | new_names == ""))

dat <- raw[-(1:4), ]
names(dat) <- new_names

if (ncol(dat) > 0 && all(grepl("^\\d+$", na.omit(as.character(dat[[1]]))))) {
  dat <- dat[, -1]
}
```

### Parse datetime

```{r parse-datetime, message=FALSE, warning=FALSE}
dt_name <- names(dat)[grepl("MM/TT/JJJ", names(dat), fixed = TRUE)][1]
if (is.na(dt_name)) stop("Datetime column not found.")

dat <- dat %>%
  mutate(datetime = mdy_hms(.data[[dt_name]], quiet = TRUE)) %>%
  filter(!is.na(datetime)) %>%
  mutate(datetime = force_tz(datetime, tz_local))
```

### Process pond data

```{r process-pond-data, message=FALSE, warning=FALSE}
pond_cols <- names(dat)[
  grepl("^[0-9]{1,2}[AB]_(0|65)_", names(dat)) &
    grepl("_(Temperature|Light)$", names(dat))
]
if (length(pond_cols) == 0) stop("No pond columns detected.")

pond_long <- dat %>%
  select(datetime, all_of(pond_cols)) %>%
  pivot_longer(-datetime, names_to = "sensor", values_to = "value") %>%
  mutate(
    unit_id  = str_match(sensor, "^([0-9]{1,2}[AB])_(0|65)_")[,2],
    depth_cm = as.numeric(str_match(sensor, "^([0-9]{1,2}[AB])_(0|65)_")[,3]),
    variable = tolower(str_match(sensor, "(Temperature|Light)$")[,2]),
    value    = suppressWarnings(as.numeric(value))
  ) %>%
  filter(variable == "temperature") %>%
  select(datetime, unit_id, depth_cm, value) %>%
  left_join(pond_map, by = "unit_id") %>%
  filter(!is.na(treatment))
```

### Process outside logger data

```{r process-outside-data, message=FALSE, warning=FALSE}
# Detect outside logger columns (Sun/Shade)
outside_cols <- names(dat)[
  grepl("^(Sun|Shade)_", names(dat)) &
    grepl("_(Temperature|Light)$", names(dat))
]

outside_long <- NULL

if (length(outside_cols) > 0) {
  outside_long <- dat %>%
    select(datetime, all_of(outside_cols)) %>%
    pivot_longer(-datetime, names_to = "sensor", values_to = "value") %>%
    mutate(
      unit_id  = str_match(sensor, "^(Sun|Shade)_")[,2],
      variable = tolower(str_match(sensor, "(Temperature|Light)$")[,2]),
      value    = suppressWarnings(as.numeric(value))
    ) %>%
    filter(!is.na(unit_id), !is.na(variable)) %>%
    select(datetime, unit_id, variable, value)
}
```

### Daily summaries

```{r daily-summaries, message=FALSE, warning=FALSE}
# Daily summaries per pond
daily_unit <- pond_long %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date, unit_id, depth_cm, treatment) %>%
  summarise(
    tmean = ifelse(all(is.na(value)), NA_real_, mean(value, na.rm = TRUE)),
    tmax  = ifelse(all(is.na(value)), NA_real_, max(value, na.rm = TRUE)),
    .groups = "drop"
  )

# Treatment daily means
daily_trt <- daily_unit %>%
  group_by(date, depth_cm, treatment) %>%
  summarise(
    tmean_av = mean(tmean, na.rm = TRUE),
    tmean_se = plotrix::std.error(tmean, na.rm = TRUE),
    tmax_av  = mean(tmax, na.rm = TRUE),
    tmax_se  = plotrix::std.error(tmax, na.rm = TRUE),
    .groups = "drop"
  )

# Outside daily means
outside_daily <- NULL

if (!is.null(outside_long)) {
  outside_daily <- outside_long %>%
    filter(variable == "temperature") %>%
    mutate(date = as.Date(datetime)) %>%
    group_by(date, unit_id) %>%
    summarise(temp = mean(value, na.rm = TRUE), .groups = "drop") %>%
    mutate(depth_cm = 0) %>%
    filter(!is.na(temp))
}
```

### Prepare plot elements

```{r prepare-plot-elements, message=FALSE, warning=FALSE}
snail_band <- tibble::tibble(
  depth_cm = 65,
  ymin = 28,
  ymax = 30,
  xmin = min(daily_trt$date, na.rm = TRUE),
  xmax = max(daily_trt$date, na.rm = TRUE)
)

# heatwave 
hw_start <- as.Date("2025-06-28")
hw_end   <- as.Date("2025-07-02")

ymax_by_depth <- daily_trt %>%
  group_by(depth_cm) %>%
  summarise(ymax = max(tmax_av, na.rm = TRUE), .groups = "drop")

heatwave_rug <- tidyr::crossing(
  depth_cm = sort(unique(daily_trt$depth_cm)),
  xmin = hw_start,
  xmax = hw_end + 1
) %>%
  left_join(ymax_by_depth, by = "depth_cm") %>%
  mutate(
    ymin = ymax - 0.6,
    ymax = ymax
  )
```

### Daily maximum temperature

```{r fig-daily-max, message=FALSE, warning=FALSE, fig.width=3.42, fig.height=4.8}
p_max <- ggplot() +
  geom_rect(
    data = heatwave_rug,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill = heatwave_col,
    alpha = 0.85
  ) +
  geom_rect(
    data = snail_band,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill = "grey70",
    alpha = 0.12
  ) +
  geom_line(
    data = daily_trt,
    aes(x = date, y = tmax_av, color = treatment),
    linewidth = 0.4
  ) +
  facet_wrap(
    ~depth_cm, ncol = 1,
    labeller = labeller(depth_cm = function(x) paste(x, "cm depth"))
  ) +
  scale_color_manual(values = cols) +
  theme_pub +
  labs(
    x = "Date",
    y = "Daily maximum temperature (°C)",
    color = "Treatment"
  )

print(p_max)

```

### Figure: Daily mean temperature (with outside loggers)

```{r fig-daily-mean, message=FALSE, warning=FALSE, fig.width=3.42, fig.height=4.8}
pA <- ggplot(daily_trt, aes(date, tmean_av, color = treatment, fill = treatment)) +
  geom_ribbon(
    aes(ymin = tmean_av - tmean_se, ymax = tmean_av + tmean_se),
    alpha = 0.25,
    color = NA
  ) +
  geom_line(linewidth = 0.4, na.rm = TRUE) +
  {if (!is.null(outside_daily)) 
    geom_line(
      data = outside_daily,
      aes(x = date, y = temp, linetype = unit_id),
      inherit.aes = FALSE,
      color = "grey40",
      linewidth = 0.5,
      alpha = 0.9,
      na.rm = TRUE
    )
  } +
  scale_linetype_manual(
    values = c(Sun = "dashed", Shade = "solid"),
    labels = c("Air (sun)", "Air (shade)")
  ) +
  facet_wrap(
    ~depth_cm,
    ncol = 1,
    labeller = labeller(depth_cm = function(x) paste(x, "cm depth"))
  ) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(
    y = "Daily mean temperature (°C)",
    x = NULL,
    color = "Treatment",
    fill  = "Treatment",
    linetype = "Outside logger"
  ) +
  theme_pub

print(pA)
```

### Effect size: Daily maximum temperature

```{r effect-size-max, message=FALSE, warning=FALSE, fig.width=3.42, fig.height=4.5}
delta_tmax <- daily_trt %>%
  select(date, depth_cm, treatment, tmax_av) %>%
  pivot_wider(names_from = treatment, values_from = tmax_av) %>%
  mutate(delta = heated - control) %>%
  filter(is.finite(delta))

med_vals <- delta_tmax %>%
  group_by(depth_cm) %>%
  summarise(med = median(delta, na.rm = TRUE), .groups = "drop")

p_delta <- ggplot(delta_tmax, aes(delta)) +
  geom_density(alpha = 0.35, fill = "firebrick") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(data = med_vals, aes(xintercept = med), linewidth = 0.7) +
  facet_wrap(
    ~depth_cm, ncol = 1,
    labeller = labeller(depth_cm = function(x) paste(x, "cm depth"))
  ) +
  theme_pub +
  labs(
    x = "Heated − Control daily maximum temperature (°C)",
    y = "Density"
  )

print(p_delta)
```

### Effect size: Daily mean temperature

```{r effect-size-mean, message=FALSE, warning=FALSE, fig.width=3.42, fig.height=4.5}
delta_tmean <- daily_trt %>%
  select(date, depth_cm, treatment, tmean_av) %>%
  pivot_wider(names_from = treatment, values_from = tmean_av) %>%
  mutate(delta = heated - control) %>%
  filter(is.finite(delta))

med_vals_mean <- delta_tmean %>%
  group_by(depth_cm) %>%
  summarise(med = median(delta, na.rm = TRUE), .groups = "drop")

label_pos_mean <- delta_tmean %>%
  group_by(depth_cm) %>%
  summarise(y = max(density(delta)$y, na.rm = TRUE), .groups = "drop") %>%
  left_join(med_vals_mean, by = "depth_cm") %>%
  mutate(
    label = paste0("Median ΔT = ", round(med, 2), " °C"),
    y = y * 0.9
  )

p_delta_mean <- ggplot(delta_tmean, aes(delta)) +
  geom_density(alpha = 0.35, fill = "firebrick") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(data = med_vals_mean, aes(xintercept = med), linewidth = 0.7) +
  geom_text(
    data = label_pos_mean,
    aes(x = med, y = y, label = label),
    inherit.aes = FALSE,
    hjust = -0.1,
    size = 4
  ) +
  facet_wrap(
    ~depth_cm, ncol = 1,
    labeller = labeller(depth_cm = function(x) paste(x, "cm depth"))
  ) +
  theme_pub +
  labs(
    x = "Heated − Control daily mean temperature (°C)",
    y = "Density"
  )

print(p_delta_mean)
```

