---
title: "Nutrients analyses"
---

## Introduction

This document analyses water quality parameters including nutrients (chloride, nitrate, phosphate, sulfate), dissolved oxygen, conductivity, pH, and water temperature under different treatments.

### Load libraries

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(plotrix)
library(stringr)
library(readr)
library(readxl)
library(patchwork)
library(lme4)
library(car)
library(tidyr)
```

### Data import

```{r data-import, message=FALSE, warning=FALSE}
df <- read.table("nutrients.txt", sep = "\t", header = TRUE)
```

### Data cleaning

```{r data-cleaning, message=FALSE, warning=FALSE}
num_cols <- c("chloride","nitrate","phosphate","sulfate","fluoride","bromide",
              "O2_pourcentage","O2_mg","conductivity","pH","water_temperature")

df2 <- df %>%
  mutate(across(all_of(num_cols), ~{
    x <- as.character(.x)
    x <- str_trim(x)
    x <- na_if(x, "")
    x <- na_if(tolower(x), "invalid")
    x <- str_replace_all(x, ",", ".")            
    x <- str_remove_all(x, "[^0-9\\.-]")        
    suppressWarnings(as.numeric(x))
  })) %>%
  mutate(
    timepoint = as.numeric(timepoint),
    treatment = factor(treatment, levels = c("control", "heated"))
  )

# Optional check
df2 %>% count(timepoint, treatment) %>% arrange(timepoint, treatment)
```

### Summary statistics

```{r summary-statistics, message=FALSE, warning=FALSE}
mdata.plot <- df2 %>%
  group_by(timepoint, treatment) %>%
  summarise(
    chloride_av  = mean(chloride, na.rm = TRUE),
    chloride_se  = plotrix::std.error(chloride, na.rm = TRUE),
    
    nitrate_av   = mean(nitrate, na.rm = TRUE),
    nitrate_se   = plotrix::std.error(nitrate, na.rm = TRUE),
    
    phosphate_av = mean(phosphate, na.rm = TRUE),
    phosphate_se = plotrix::std.error(phosphate, na.rm = TRUE),
    
    sulfate_av   = mean(sulfate, na.rm = TRUE),
    sulfate_se   = plotrix::std.error(sulfate, na.rm = TRUE),
    
    fluoride_av  = mean(fluoride, na.rm = TRUE),
    fluoride_se  = plotrix::std.error(fluoride, na.rm = TRUE),
    
    bromide_av   = mean(bromide, na.rm = TRUE),
    bromide_se   = plotrix::std.error(bromide, na.rm = TRUE),
    
    O2_pourcentage_av = mean(O2_pourcentage, na.rm = TRUE),
    O2_pourcentage_se = plotrix::std.error(O2_pourcentage, na.rm = TRUE),
    
    O2_mg_av     = mean(O2_mg, na.rm = TRUE),
    O2_mg_se     = plotrix::std.error(O2_mg, na.rm = TRUE),
    
    conductivity_av = mean(conductivity, na.rm = TRUE),
    conductivity_se = plotrix::std.error(conductivity, na.rm = TRUE),
    
    pH_av        = mean(pH, na.rm = TRUE),
    pH_se        = plotrix::std.error(pH, na.rm = TRUE),
    
    water_temp_av = mean(water_temperature, na.rm = TRUE),
    water_temp_se = plotrix::std.error(water_temperature, na.rm = TRUE),
    
    .groups = "drop"
  )

tp_levels <- sort(unique(df2$timepoint))
mdata.plot <- mdata.plot %>%
  mutate(time_f = factor(timepoint, levels = tp_levels))
```

### Plot settings

```{r plot-settings, message=FALSE, warning=FALSE}
my_scale <- scale_color_manual(
  name   = "Treatment",
  values = c("control" = "#0072B2", "heated" = "#D55E00"),
  labels = c("Control", "Heated")
)

nut_theme <- theme_classic(base_size = 16) +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    axis.title = element_text(face = "bold", size = 16),
    axis.text  = element_text(size = 14, color = "black"),
    axis.title.x = element_text(face = "bold", size = 16, color = "black"),
    axis.title.y = element_text(face = "bold", size = 16, color = "black"),
    legend.position = "none",
    legend.title = element_blank(),
    legend.text  = element_text(size = 14),
    plot.title   = element_text(hjust = 0.5)
  )

x_scale <- scale_x_discrete(drop = FALSE)

make_plot <- function(dat, y, se, ylab, ylim = NULL, xlab = NULL) {
  
  p <- ggplot(dat, aes(x = time_f, y = .data[[y]], color = treatment, group = treatment)) +
    nut_theme +
    geom_pointrange(
      aes(ymin = .data[[y]] - .data[[se]], ymax = .data[[y]] + .data[[se]]),
      position = position_dodge(width = 0.35),
      na.rm = TRUE
    ) +
    geom_line(linewidth = 1, na.rm = TRUE) +
    my_scale +
    x_scale +
    labs(x = xlab, y = ylab)
  
  if (!is.null(ylim)) {
    p <- p + coord_cartesian(ylim = ylim)
  }
  
  p
}
```

### Individual plots

```{r individual-plots, message=FALSE, warning=FALSE}
px_chloride     <- make_plot(mdata.plot, "chloride_av", "chloride_se", "Chloride concentration", ylim = c(50, NA))
px_nitrate      <- make_plot(mdata.plot, "nitrate_av", "nitrate_se", "Nitrate concentration")
px_phosphate    <- make_plot(mdata.plot, "phosphate_av", "phosphate_se", "Phosphate concentration")
px_sulfate      <- make_plot(mdata.plot, "sulfate_av", "sulfate_se", "Sulfate concentration")
px_O2_pct       <- make_plot(mdata.plot, "O2_pourcentage_av", "O2_pourcentage_se", "O2 (%)")
px_conductivity <- make_plot(mdata.plot, "conductivity_av", "conductivity_se", "Conductivity")
px_pH           <- make_plot(mdata.plot, "pH_av", "pH_se", "pH")

# Water temperature: more detailed y axis
ymin_temp <- floor(min(mdata.plot$water_temp_av - mdata.plot$water_temp_se, na.rm = TRUE))
ymax_temp <- ceiling(max(mdata.plot$water_temp_av + mdata.plot$water_temp_se, na.rm = TRUE))

px_water_temp <- make_plot(mdata.plot, "water_temp_av", "water_temp_se", "Water temperature", xlab = "Weeks") +
  scale_y_continuous(
    breaks = seq(ymin_temp, ymax_temp, by = 2),
    expand = c(0, 0)
  )
```


### Statistical analyses

```{r prepare-filtered-data, message=FALSE, warning=FALSE}
# Prepare filtered data (exclude timepoints 0 and 1)
num_cols_stats <- c("chloride", "nitrate", "phosphate", "sulfate",
                    "O2_pourcentage", "water_temperature", "conductivity", "pH")

df_filtered <- df2 %>% filter(!timepoint %in% c(0, 1)) %>% mutate(pond = factor(pond), block = factor(block))
```

#### Water temperature

```{r stats-temperature, message=FALSE, warning=FALSE}
model_temp <- lmer(water_temperature ~ treatment * timepoint + (1 | pond) + (1 | block), data = df_filtered)
Anova(model_temp, test = "F")
summary(model_temp)
```

#### Chloride

```{r stats-chloride, message=FALSE, warning=FALSE}
model_chloride <- lmer(chloride ~ treatment * timepoint + (1 | pond) + (1 | block), data = df_filtered)
Anova(model_chloride, test = "F")
summary(model_chloride)
```

#### Nitrate

```{r stats-nitrate, message=FALSE, warning=FALSE}
model_nitrate <- lmer(nitrate ~ treatment * timepoint + (1 | pond) + (1 | block), data = df_filtered)
Anova(model_nitrate, test = "F")
summary(model_nitrate)
```

#### Phosphate

```{r stats-phosphate, message=FALSE, warning=FALSE}
model_phosphate <- lmer(phosphate ~ treatment * timepoint + (1 | pond) + (1 | block), data = df_filtered)
Anova(model_phosphate, test = "F")
summary(model_phosphate)
```

#### Sulfate

```{r stats-sulfate, message=FALSE, warning=FALSE}
model_sulfate <- lmer(sulfate ~ treatment * timepoint + (1 | pond) + (1 | block), data = df_filtered)
Anova(model_sulfate, test = "F")
summary(model_sulfate)
```

#### Dissolved oxygen

```{r stats-oxygen, message=FALSE, warning=FALSE}
model_O2 <- lmer(O2_pourcentage ~ treatment * timepoint + (1 | pond) + (1 | block), data = df_filtered)
Anova(model_O2, test = "F")
summary(model_O2)
```

#### Conductivity

```{r stats-conductivity, message=FALSE, warning=FALSE}
model_cond <- lmer(conductivity ~ treatment * timepoint + (1 | pond) + (1 | block), data = df_filtered)
Anova(model_cond, test = "F")
summary(model_cond)
```

#### pH

```{r stats-ph, message=FALSE, warning=FALSE}
model_ph <- lmer(pH ~ treatment * timepoint + (1 | pond) + (1 | block), data = df_filtered)
Anova(model_ph, test = "F")
summary(model_ph)
```