---
title: "Indoor experiment: Growth, herbivory, and snail mortality"
---

## Introduction

This document analyses the indoor experiment data examining duckweed growth, herbivory rates, and snail mortality across different temperatures.

### Load libraries

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(patchwork)
library(lme4)
library(car)
```

### Data import and preparation

```{r data-import, message=FALSE, warning=FALSE}
dat <- read.table("sn_duck_ind.txt", sep = "\t", header = TRUE)

dat <- dat %>%
  mutate(
    temperature = as.numeric(temperature),
    frond_change_day = fronds_end - fronds_start
  )
```

### Calculate control baseline and herbivory metrics

```{r calculate-metrics, message=FALSE, warning=FALSE}
control_baseline <- dat %>%
  filter(treatment == "control") %>%
  group_by(round, temperature) %>%
  summarise(
    control_mean_change = mean(frond_change_day, na.rm = TRUE),
    .groups = "drop"
  )
dat2 <- dat %>%
  left_join(control_baseline, by = c("round", "temperature")) %>%
  mutate(
    corrected_herbivory = if_else(
      treatment == "herbivory",
      control_mean_change - frond_change_day,
      NA_real_
    ),
    snail_effective = (snail_start + snail_end) / 2,
    corrected_herbivory_per_snail = if_else(
      treatment == "herbivory",
      corrected_herbivory / snail_effective,
      NA_real_
    )
  )

temps <- sort(unique(dat2$temperature))
```

### Summary statistics

#### Herbivory summary

```{r herbivory-summary, message=FALSE, warning=FALSE}
herb <- dat2 %>% filter(treatment == "herbivory")

round_means_snail <- herb %>%
  group_by(round, temperature) %>%
  summarise(
    mean_corr_snail = mean(corrected_herbivory_per_snail, na.rm = TRUE),
    .groups = "drop"
  )

temp_summary_snail <- round_means_snail %>%
  group_by(temperature) %>%
  summarise(
    mean = mean(mean_corr_snail, na.rm = TRUE),
    se   = sd(mean_corr_snail, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(series = "Herbivory")
```

#### Growth summary

```{r growth-summary, message=FALSE, warning=FALSE}
# Growth (control)
growth_dat <- dat2 %>%
  filter(treatment == "control") %>%
  mutate(
    growth_rate    = fronds_end - fronds_start,     # fronds per day
    temperature    = as.numeric(temperature),       # for plotting on x-axis
    temperature_f  = factor(temperature, levels = sort(unique(temperature)))  # for modeling
  )

# Mean growth per round × temperature 
round_growth <- growth_dat %>%
  group_by(round, temperature) %>%
  summarise(
    mean_growth = mean(growth_rate, na.rm = TRUE),
    .groups = "drop"
  )

# Mean ± SE across rounds (one value per temperature)
temp_growth_summary <- round_growth %>%
  group_by(temperature) %>%
  summarise(
    mean = mean(mean_growth, na.rm = TRUE),
    se   = sd(mean_growth, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(series = "Growth")
```

#### Mortality summary

```{r mortality-summary, message=FALSE, warning=FALSE}
# Mortality (bars): mean across rounds
mortality_by_round <- herb %>%
  group_by(round, temperature) %>%
  summarise(
    deaths         = sum(snail_start - snail_end, na.rm = TRUE),
    total_snails   = sum(snail_start, na.rm = TRUE),
    mortality_rate = deaths / total_snails,
    .groups = "drop"
  )

mortality_mean <- mortality_by_round %>%
  group_by(temperature) %>%
  summarise(
    mean_mortality = mean(mortality_rate, na.rm = TRUE),
    .groups = "drop"
  )
```

### Plot settings

```{r plot-settings, message=FALSE, warning=FALSE}
growth_color <- "#006D6F"   # deep teal
herb_color   <- "#B03A2E"   # warm brick
mort_fill    <- "grey80"
mort_edge    <- "grey30"

axis_lw  <- 0.55
trend_lw <- 1.20
err_lw   <- 0.75

base_theme <- theme_classic(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = axis_lw, color = "black"),
    axis.ticks = element_line(linewidth = axis_lw, color = "black"),
    axis.ticks.length = unit(2.2, "mm"),
    legend.title = element_blank(),
    legend.key = element_rect(fill = "white", color = NA)
  )
```

### Growth and herbivory rates

```{r panel-a, message=FALSE, warning=FALSE, fig.width=3.4, fig.height=4}
y_breaks <- seq(0, 30, by = 5)
y_minor  <- seq(0, 30, by = 2.5)

p_top <- ggplot() +
  # faint round points (behind)
  geom_point(
    data = round_df,
    aes(x = temperature, y = value, color = series),
    alpha = 0.18,
    size = 1.6,
    shape = 16,
    show.legend = FALSE
  ) +
  # mean ± SE
  geom_errorbar(
    data = means_df,
    aes(x = temperature, ymin = mean - se, ymax = mean + se, color = series),
    width = 0,
    linewidth = err_lw
  ) +
  geom_line(
    data = means_df,
    aes(x = temperature, y = mean, color = series, group = series),
    linewidth = trend_lw
  ) +
  geom_point(
    data = means_df,
    aes(x = temperature, y = mean, color = series),
    size = 3.7
  ) +
  scale_color_manual(values = c("Growth" = growth_color, "Herbivory" = herb_color)) +
  scale_x_continuous(breaks = temps) +
  coord_cartesian(ylim = c(2.5, 28.5)) +
  scale_y_continuous(
    breaks = y_breaks,
    minor_breaks = y_minor,
    expand = expansion(mult = c(0.02, 0.04))
  ) +
  labs(
    x = NULL,
    y = expression("Rate (fronds" %.% "day"^{-1}*")")
  ) +
  base_theme +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x  = element_blank(),
    
    legend.position = c(0.22, 0.86),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.text = element_text(size = 10),
    guides(color = guide_legend(override.aes = list(size = 2.8))),
    
    plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.margin = margin(6, 10, 0, 10)
  ) +
  labs(title = "A")

p_top
```

### Snail mortality

```{r panel-b, message=FALSE, warning=FALSE, fig.width=3.4, fig.height=2.2}
p_bottom <- ggplot(mortality_mean, aes(x = temperature, y = mean_mortality)) +
  geom_col(
    fill = mort_fill,
    color = mort_edge,
    linewidth = 0.5,
    width = 0.90
  ) +
  scale_x_continuous(breaks = temps) +
  coord_cartesian(ylim = c(0, 0.10)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  labs(
    x = expression("Temperature ("*degree*"C)"),
    y = "Snail mortality (proportion dead)"
  ) +
  base_theme +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.margin = margin(0, 10, 6, 10)
  ) 
p_bottom
```

### Statistical analyses

#### Herbivory analysis

```{r stats-herbivory, message=FALSE, warning=FALSE}
herb_model <- dat2 %>% filter(treatment == "herbivory")
herb_model$temperature <- factor(herb_model$temperature)

# Mixed model
m1 <- lmer(corrected_herbivory_per_snail ~ temperature + (1 | round), data = herb_model)
Anova(m1, test = "F")
summary(m1)
```

#### Growth analysis

```{r stats-growth, message=FALSE, warning=FALSE}
# Mixed model for growth
m2 <- lmer(growth_rate ~ temperature_f + (1 | round), data = growth_dat)
Anova(m2, test = "F")
summary(m2)
```